How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-07-08 doc/response_review.tex

  The caveat that human impacts may dramatically alter extinction risk seems like an important one and, as it stands, seems like it could undercut the (rather terse) final conclusion. Would conservation decisions really be bolstered by including fossil data? I would agree with that (and of course I would, as a paleobiologist), but it might be useful to expand a bit on this topic. Do you mean that a model such as yours, incorporating geographic range and other such parameters, would be helpful? In what way could it be used? Or do you mean using models based on past intervals with environmental changes inferred to be similar – the PETM as an analogue for warming and acidification, for example? I wonder if it would be helpful to draw a stronger connection linking your finding to conservation biology, given the different stressors now.
\end{bfseries}

WAITING FOR SETH

\begin{bfseries}
  I found the description on lines 171-174 to be a bit confusing. The phrase “this pattern” (line 171) seems to refer to the posterior distribution less than/equal to 0.5. The sentence states that the pattern is absent in absent in VP model for forams and radiolarians, but to me the radiolarian pattern doesn’t look too different from the V model (which also doesn’t dip below 0.5). The next sentence argue that there are fewer period of low performance for calcareous nannos and dinoflagellates in the VP model, but the differences seem minimal. Perhaps I’m not familiar with these graphs, but there only seems to be a difference in calcareous nannos around 50 Ma, and among dinoflagellates a small difference around 58 Ma and 30 Ma?
\end{bfseries}

Thank you for pointing out this lack of clarity in the Results section. We have revised this section for clarity.

\begin{bfseries}
  Figures 2-4. I guessed (and later confirmed) that the alternating gray and white bars represented stages of the geological time scale, but it would be helpful to indicate that in the captions.
\end{bfseries}

We have updated the relevant sections of the text with these suggestions.

\subsection{Reviewer 3}

\begin{bfseries}
  This is an interesting paper on extinction prediction using microfossil plankton data from deep-sea cores.

  It’s very important that they found simple model with only a few parameters of geographic range, temperature etc can robustly predict extinction risk regardless of minor model difference.

  I enjoyed my reading and hope the comments below help to improve the ms.

  1. just in case, for foraminifera, benthic species are deleted from the database? Or they originally include planktonic foram only?
\end{bfseries}

We have confirmed using MicroTax that all foram species included our data set our planktonic species only.

\begin{bfseries}
  2. Neptune database has an internally consistent taxonomic identification strategy. Is this true? I guess it’s non-critical compilation of census data done by shipboard micropaleontologists over ~50 years (ie taxonomic sense of different generation micropaleontologists are very different). David Lazarus (in charge person of Neptune) is radiolarian person, and so radiolarian may be better in taxonomy? Planktonic foram diversity is low, so may be better straightforward. Anyway, I recommend deeper reference search and reading to see how is Neptune taxonomy and explain it in the method section more in details.
\end{bfseries}

We have added a statement in our introduction discussing that the Neptune database has an internal synonymization framework as detailed in \citet{Lazarus1994}. While the taxonomic identities may not be the most accepted by the community, they are consistent within the database.

\begin{bfseries}
  3. Fig 1, better to explain what are C, V, CP, VP, AUC in the caption

  4. Fig 2 and 3, What do gray bars mean? Gray bay = envelope?

  5. It will help our understanding if you explain the main point/message of each fig in the contest of discussion in each caption 
\end{bfseries}

Thank you for these suggestions. We have updated our figure captions to be more descriptive. Additionally, we have clarified the link between our figure elements and our model description table.

\begin{bfseries}
  6. I am not sure if it’s possible, but I am curious where geographic range or temperature are the key? Or both are essential? It is possible to try geographic range only model and temperature only model to see such?
\end{bfseries}

Thank you for your suggestion. We chose to always included geographic range and temperature because these represent basic measurements of a species and its environmental context. Additionally, if we were to use models with one of these covariates and not the other, then we are assuming that the ignored covariate is known with 100\% probability of having 0 effect on extinction risk. We are uncomfortable making this assumption and instead Instead we use weakly informative priors to limit spuriously strong parameter estimates and ``shrink'' effects towards 0 if that covariate has a weak outcome on the response.

We have also added a section to the supplement with the group-level and individual-level estimates for all of our regression coefficients to demonstrate the size of our covariate effects.

\begin{bfseries}
  7. The authors use deep-water temperature (using geochemistry of benthic foram from deep-sea cores) data. And their biotic data is plankton. Their plankton can be mostly from ocean surface or photic zone (eg diatoms), but some have good bunch of deep-water species (radioralians).

  The authors need some discussion and justification on this.
\end{bfseries}

We have added a section to the Methods discussing this data limitation

From the updated manuscript:

``Mg/Ca based temperature estimates are measured from benthic forams, and are an estimate of deep water ocean temperature. The organisms in this study are all planktonic, Mg/Ca based temperature estimates do not describe the exact environment these organisms inhabit. Ideally, we would have detailed ocean surface water temperature estimates for the entire globe for the entire Cenozoic. Unfortunately, that type of data does not exist. So, we interpret our temperature estimates as reflecting the global climate state that an organisms experiences, and not as a descriptor of that taxon's local environmental ecology.''

\begin{bfseries}
  8. While geographical range consider spatial stuff, the authors use one temperature curve for all (I understand there will be no other way). Is there any difference in temperature sensitivity between narrow and wide distribution species; or low latitude and high latitude species? I understand this may be out of the scope. The authors may ignore.
\end{bfseries}

We thank the reviewer for this comment. We do agree that surface temperature estimates, especially geographically localized ones, would be a major improvement to our data. However, detailed sea-surface estimates for the entire Cenozoic are not readily available and generating those would be outside the scope of this study.

\printbibliography
\end{refsection}

\end{document}
